%% Modified documentstyle to documentclass -- compatibility with LaTex2e
%% N. Mancell (98/03/01).

%%  ucalgthes_root.tex        (NM 98/03/01)
%   Modified   92-09-18      Add references to dissertation        D. Teale
%                            Add approval page to toc
%                            Add ref to Title Degree on approval page
%   Modified  2006-09-12     Added geometry package to set up UofC thesis margins
%                            Removed includeprompt option N. Mancell               
%
\documentclass{ucalgthes1}   
\usepackage[letterpaper,top=1in, bottom=1.22in, left=1.40in, right=0.850in]{geometry}
\usepackage{fancyhdr}
\fancyhead{}
\fancyfoot{}
\renewcommand{\headrulewidth}{0pt}
\fancyhead[RO,LE]{\thepage}  
%Define other usepackages here
\usepackage{hyperref}
\usepackage{MyDecls}
\title{What is a Differential Partial Combinatory Algebra?}
%
%            Insert the correct information between the {}
%
\author{Jonathan Gallagher}
\thesisyear{2011}
\thesis{thesis}    % the word dissertation can be inserted between {}
\newcommand{\thesistitle}{What is a Differential Partial Combinatory Algebra?}
\monthname{January}
\dept{Computer Science}
\degree{Master of Science}
%
%                    End of supplied information
%
\begin{document}
\makethesistitle
\pagenumbering{roman}     % resets page counter to one
\setcounter{page}{1}
\chapter*{UNIVERSITY OF CALGARY \\ FACULTY OF GRADUATE STUDIES}
\thispagestyle{empty}
The undersigned certify that they have read, and recommend
to the Faculty of Graduate Studies for acceptance, a \Thesis\ entitled
``\thesistitle'' submitted by \Author\
in partial fulfillment of the requirements for the degree of
\Degree.\\

%
%                 Substitute  List of Examiners
%
\begin{signing}{Department of Academic Computing}
\signline
Supervisor, Dr.~J.~ Robin Cockett \\
Department of Computer Science\\
\signline
Dr.~Clifton Cunningham \\
Department of Mathematics and Statistics\\
\signline
Dr. Michael J. Jacobson \\
Department of Computer Science\\
\end{signing}
%Differential Structure in Computability Theory \\ \bigskip Second ThesisTitle Line
\newpage
\phantomsection
\altchapter{Abstract} 
In this thesis we combine Turing categories with Cartesian left additive restriction categories and again with differential restriction categories.  The result of the first combination is a new structure which models nondeterministic computation.  The result of the second combination is a structure which models the notion of linear resource consumption.  We also study the structural background required to understand what new features Turing structure should have in light of addition and differentiation -- most crucial to this development is the way in which idempotents split.  For the combination of Turing categories with Cartesian left additive restriction categories we will also provide a model.
\newpage
\phantomsection
\altchapter{Acknowledgements}
For this thesis, my deepest thanks and gratitude goes to Professor Robin Cockett.  From finding opportunities for me to speak at conferences and workshops to helping me make my ideas into coherent thoughts to giving me a push back on track when I had failed -- prof. Cockett has played a very important role in my development as a graduate student.


Next, I want to thank Dr. Geoff Cruttwell ``getting me going'' in my studies, and helping me with writing.


I am grateful to Professor Clifton Cunningham and Professor Michael Jacobson for serving on my thesis committee.


Two special ``thank you's'' to Professor Pieter Hofstra and Professor Peter Selinger.  Professor Hofstra supported me to attend and speak at the Semigroups and Categories workshop at the University of Ottawa, and Professor Selinger supported me to attend and speak at the Octoberfest category theory workshop in Dalhousie University.


\begin{singlespace}
\newpage
\phantomsection
\tableofcontents
\pagestyle{plain}
\newpage
\phantomsection
\listoftables
\pagestyle{plain}
\newpage
\phantomsection
% \listoffigures
\pagestyle{plain}
\clearpage
\end{singlespace}
\clearpage          % otherwise tables will be numbered wrong

\pagenumbering{arabic}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BEGIN THESIS CONTENT %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\include{chapter0}
\include{chapter1} %Cartesian Restriction; Simple Slice; Idempotent Splitting; Term Logic; Turing
\include{chapter2} %Cartesian Left additive restriction; Simple Slice; Idempotent Splitting; Term Logic; Turing
\include{chapter3} %Differential restriction; Simple Slice; Idempotent Splitting; Term Logic; Turing
% \include{chapter5} % Cokleisli
% \include{chapter4} % Rat
% \include{TermLog} % Term logic
% \appendix
% \include{appendix1}

%% Bibliography
\bibliographystyle{plain}
\bibliography{../CitationDBs/References}
%\bibliography{/home/jdgallag/SVNLink/papers/JonathanGallagher/CitationDBs/References}

\end{document}
